Master Thesis

### 2013-06-13 acg.tex
Almost finished writing about defining signatures
Finished writing 'Defining signatures'
Prettier implementation of lexicons
Switched to the report class, start using theorem envs

\subsection{Abstract Categorial Grammars}

@@

\subsubsection{Higher-order signatures}

@@

\subsubsection{Example signature}
\label{sssec:example-sig}

@@

\subsubsection{Lexicons}

@@

\subsubsection{Example lexicon}
\label{sssec:example-lex}

@@

To illustrate the ideas of a lexicon and an abstract categorial grammar,
we will expand our example from \ref{sssec:example-sig}. We will
consider another signature, $\Sigma_{Synt}$, which will describe the
syntax of quantified noun phrases in the style of Montague
\cite{montague1973proper}. The atomic types $A_{Synt}$ will consist of
the types $np$, $n$ and $s$. Our constants, $C_{Synt} = \{\synt{every},
\synt{some}, \synt{love}, \synt{man}, \synt{woman}\}$, will have the
following types, as predicted by $\tau_{Synt}$:
\begin{itemize}
\item $\tau_{Synt}(\synt{every}) = \tau_{Synt}(\synt{some}) = n \limp
  ((np \limp s) \limp s)$
\item $\tau_{Synt}(\synt{love}) = np \limp np \limp s$
\item $\tau_{Synt}(\synt{man}) = \tau_{Synt}(\synt{woman}) = n$
\end{itemize}

@@

\subsubsection{Example semantic lexicon}

One perk of working with abstract categorial grammars is that one
abstract term can be interpreted multiple ways, using different
lexicons. In the previous section, we considered the abstract terms of
the $\Sigma_{Synt}$ signature and interpreted them as computations on
strings. Now we will take the same abstract terms, but try to intepret
them as computations on entities and truth values.

@@

Now to see how this signature connects to our abstract signature
$\Sigma_{Synt}$, we will give a lexicon from the latter to the
former. First, we give the type mappings. $\mathcal{L}_{sem}(n) = e
\limp t$ meaning that nouns will yield boolean functions of one argument
(unary \emph{predicates} in functional programming
terminology). $\mathcal{L}_{sem}(np) = e$, simple noun phrases will be
mapped into terms which yield some entity. Finally,
$\mathcal{L}_{sem}(s) = t$, sentences will be assigned terms which
compute their truthiness. Let us see how this plan is achieved in the
$\mathcal{L}_{sem}$ lexicon.
\begin{itemize}
\item $\mathcal{L}_{sem}(\synt{every}) = \lambda^{\circ} P Q.\ \forall
  x.\ ((P\ x) \Rightarrow (Q\ x))$\footnote{We will be using $\forall
    x.\ M$ as a shortcut for $\forall\ (\lambda x.\ M)$ and $\exists
    x.\ M$ for $\exists\ (\lambda x.\ M)$.}
\item $\mathcal{L}_{sem}(\synt{some}) = \lambda^{\circ} P Q.\ \exists
  x.\ ((P\ x) \land (Q\ x))$
\item $\mathcal{L}_{sem}(\synt{love}) = \textbf{love}$
\item $\mathcal{L}_{sem}(\synt{man}) = \textbf{man}$
\item $\mathcal{L}_{sem}(\synt{woman}) = \textbf{woman}$
\end{itemize}

Let us now consider the example terms of the $\Sigma_{Synt}$ signature
from \ref{sssec:example-lex}. $t_3 = \synt{some}\ \synt{woman}$ will be
interpreted as $\mathcal{L}_{sem}(t_3) =_{\beta} \lambda^{\circ}
Q.\ \exists x.\ ((\textbf{woman}\ x) \land (Q\ x))$ having type $(e \limp
t) \limp t$. It is a function that expects some predicate function and
tests whether that predicate holds at least for some woman (some entity
for which the $\textbf{woman}$ predicate holds).

>>>>

\section{Abstract Categorial Grammars}
\label{sec:acg}

@@

\subsection{Higher-order signatures}
\label{ssec:sig}

@@

\subsection{Example signature}
\label{ssec:example-sig}

@@

\subsection{Lexicons}

@@

\subsection{Example lexicon}
\label{ssec:example-lex}

@@

To illustrate the ideas of a lexicon and an abstract categorial grammar,
we will expand our example from \ref{ssec:example-sig}. We will
consider another signature, $\Sigma_{Synt}$, which will describe the
syntax of quantified noun phrases in the style of Montague
\cite{montague1973proper}. The atomic types $A_{Synt}$ will consist of
the types $np$, $n$ and $s$. Our constants, $C_{Synt} = \{\synt{every},
\synt{some}, \synt{love}, \synt{man}, \synt{woman}\}$, will have the
following types, as predicted by $\tau_{Synt}$:
\begin{itemize}
\item $\tau_{Synt}(\synt{every}) = \tau_{Synt}(\synt{some}) = n \limp
  ((np \limp s) \limp s)$
\item $\tau_{Synt}(\synt{love}) = np \limp np \limp s$
\item $\tau_{Synt}(\synt{man}) = \tau_{Synt}(\synt{woman}) = n$
\end{itemize}

@@

\subsection{Example semantic lexicon}
\label{ssec:acg-sem}

One perk of working with abstract categorial grammars is that one
abstract term can be interpreted multiple ways, using different
lexicons. In the previous chapter, we considered the abstract terms of
the $\Sigma_{Synt}$ signature and interpreted them as computations on
strings. Now we will take the same abstract terms, but try to intepret
them as computations on entities and truth values.

@@

Now to see how this signature connects to our abstract signature
$\Sigma_{Synt}$, we will give a lexicon from the latter to the
former. First, we give the type mappings. $\mathcal{L}_{sem}(n) = e
\limp t$ meaning that nouns will yield boolean functions of one argument
(unary \emph{predicates} in functional programming
terminology). $\mathcal{L}_{sem}(np) = e$, simple noun phrases will be
mapped into terms which yield some entity. Finally,
$\mathcal{L}_{sem}(s) = t$, sentences will be assigned terms which
compute their truthiness. Let us see how this plan is achieved in the
$\mathcal{L}_{sem}$ lexicon.
\begin{itemize}
\item $\mathcal{L}_{sem}(\synt{every}) = \lambda^{\circ} P Q.\ \forall
  x.\ ((P\ x) \Rightarrow (Q\ x))$\footnote{We will be using $\forall
    x.\ M$ as a shortcut for $\forall\ (\lambda x.\ M)$ and $\exists
    x.\ M$ for $\exists\ (\lambda x.\ M)$.}
\item $\mathcal{L}_{sem}(\synt{some}) = \lambda^{\circ} P Q.\ \exists
  x.\ ((P\ x) \land (Q\ x))$
\item $\mathcal{L}_{sem}(\synt{love}) = \lambda^{\circ} s o. (\textbf{love}\ s\ o)$
\item $\mathcal{L}_{sem}(\synt{man}) = \lambda^{\circ} x. (\textbf{man}\ x)$
\item $\mathcal{L}_{sem}(\synt{woman}) = \lambda^{\circ} x. (\textbf{woman}\ x)$
\end{itemize}

Let us now consider the example terms of the $\Sigma_{Synt}$ signature
from \ref{ssec:example-lex}. $t_3 = \synt{some}\ \synt{woman}$ will be
interpreted as $\mathcal{L}_{sem}(t_3) =_{\beta} \lambda^{\circ}
Q.\ \exists x.\ ((\textbf{woman}\ x) \land (Q\ x))$ having type $(e \limp
t) \limp t$. It is a function that expects some predicate function and
tests whether that predicate holds at least for some woman (some entity
for which the $\textbf{woman}$ predicate holds).

### 2013-06-13 ig.tex
Talked about the 3 mechs and their naive combo
Switched to the report class, start using theorem envs

\subsection{Interaction Grammars}

@@

\subsubsection{Frigram, an interaction grammar}

The reason we are interested in interaction grammars is because of the
existence of Frigram, a large scale grammar of French which is
lexicalized by Frilex, an independent lexical resource that we will be
building our grammar on as well. Furthermore, as we will see
in~\ref{sssec:link-ig-acg}, interaction grammars are closely linked with
abstract categorial grammars, which makes Frigram a suitable grammar to
use as a guide when developing our grammar.

In this section, we will introduce the metagrammatical structure of
Frigram and talk about how it uses the formalism of interaction grammars
to solve several tricky linguistic phenomena.

@@

\subsubsection{The link with abstract categorial grammars}
\label{sssec:link-ig-acg}

>>>>

\section{Interaction Grammars}

@@

\subsection{Frigram, an interaction grammar}
\label{ssec:frigram}

The reason we are interested in interaction grammars is because of the
existence of Frigram, a large scale grammar of French which is
lexicalized by Frilex, an independent lexical resource that we will be
building our grammar on as well. Furthermore, as we will see
in~\ref{ssec:link-ig-acg}, interaction grammars are closely linked with
abstract categorial grammars, which makes Frigram a suitable grammar to
use as a guide when developing our grammar.

In this chapter, we will introduce the metagrammatical structure of
Frigram and talk about how it uses the formalism of interaction grammars
to solve several tricky linguistic phenomena.

@@

\subsection{The link with abstract categorial grammars}
\label{ssec:link-ig-acg}

### 2013-06-13 my-work-rewritten.tex
Switched to the report class, start using theorem envs

\section{Wide-coverage abstract categorial grammars}

@@

\subsection{The tools and techniques}

@@

\subsubsection{The case for relational programming}

@@

Finally, some of the structures that we will encounter in our domain are
more directly modelled in mathematics as a relation instead of a
function. This has the benefit that to implement a relation in
relational programming, one just writes the relation, while if we were
forced to encode everything in functions, we would have to manage this
multiplicity of values manually. Examples of structures that have this
kind of structures if Frilex, our lexical database, where a single
wordform can be mapped to multiple hypertags. We will see another
example of a relation within our domain in the next subsection and we
will also discuss the advantages and drawbacks of modelling functions as
relations.

\subsubsection{Choice of implementation language}

@@

\subsection{Defining signatures and lexicons}

\subsubsection{Defining signatures}

In~\ref{sssec:sig}, we formally presented signatures as triples
$\mathopen{<}A, C, \tau\mathclose{>}$. The set $C$ can be easily
reconstructed as the domain of $\tau$ and the set $A$ is just the set of
all atomic types occurring in the range of $\tau$. Therefore, to define
the signature $\mathopen{<}A, C, \tau\mathclose{>}$, it is enough to
provide the function $\tau$, which in its set-theoretical realization is
just a set of pairs. In relational programming, a set can be easily
encoded as a unary relation that answers the membership test. Therefore,
our goal will be to construct to answer this relation and this relation
will be our final representation of a signature.

@@

\begin{description}
  \item[Function of wordform] We could imagine that every different
    wordform of the lexical database will end up being mapped to some
    constant in the signature. However, this is unsuitable since the
    lexical database is a relation that can assign more than one
    distinct hypertag to a single wordform (e.g. with different part of
    speech) and we would like to have constants of different types for
    each of these hypertags.
  \item[Function of hypertag] Conversely, we might imagine that the
    hypertags of the lexical database are mapped via a function to the
    constants of the signature. Here we run into trouble when we
    consider variations in spelling or phonology. We might have two
    items in our lexical database that share the same hypertag but whose
    wordforms still differ. We would like to have two distinct constants
    in our signature then, so that when these two constants are mapped
    by the surface lexicon, they will yield two different strings.
  \item[Function of lexical entry] Since neither the wordform nor the
    hypertag is enough, we could consider the entire lexical entry (an
    element of the lexical database, a pair of a wordform and one of its
    hypertags). This solves both of the above problems but it is still
    not perfect. Consider the case of a determiner and some semantic
    signature. Since our link between elements of the lexical database
    and of our signature is a function, there must be a constant in the
    signature for every item of our lexical database. If you recall
    in~\ref{sssec:acg-sem}, the constants of the semantic signature
    $\Sigma_{Sem}$ contained predicates for the common nouns $man$ and
    $woman$ and for the verb $loves$ and some common logical
    furniture. However, it did not contain any constants which would
    correspond to the determiners $every$ and $some$.
  \item[Partial function of lexical entry] We could fix the issue above
    by using a partial function. This accounts for cases when a lexical
    entry has no constants to represent it in a signature. Nevertheless,
    this only fixes half of the problem. We will consider the single
    lexical entry which represents e.g. the 3rd person singular present
    tense form $loves$ of the transitive verb $love$. A legitimate
    account of subject/object scope ambiguity would be to consider two
    constants in our abstract syntactic signature that have distinct
    images in some semantic lexicon, one giving the outer scope to the
    subject and the other to the object. This means that in order to
    define such a signature, we would need to link a single lexical
    entry in our database to more than one constant in the signature.
  \item[Relation between lexical entry and constant] Finally,
    considering all the above common scenarios, the only mathematical
    model general enough to handle all of the above cases is that of a
    relation. In our system, this will correspond exactly to a relation
    that will be provided by the grammar author.
\end{description}

@@

The system takes care of wiring up the relations and assembling the
signature. The grammar author's duty is to provide a 4-ary relation that
links the four parts of a lexical constant: the wordform, the hypertag,
the specifier and the type. This relation can be thought of as a
multi-valued ``function'' that receives as input a wordform and a
hypertag (that is, a single lexical entry) and outputs (possibly
multiple) pairs of specifiers and types. Since we are working in a
functional programming language that supports manipulation of
higher-order functions, we can provide helper functions which construct
these 4-ary relations for the grammar author in some simple cases. Our
library thus provides functions like \texttt{unitypedr}, which expects a
type and returns the 4-ary relation that assigns to every lexical item a
single constant whose specifier is $nil$ and whose type is the supplied
type, or \texttt{ht->typer}, which expects a mapping from hypertag
patterns to types and returns the 4-ary relation which tries to map the
hypertag against the patterns and yields constants with the
corresponding types and $nil$ specifiers. When other patterns emerge,
the grammar author is completely free to implement her new abstractions
using the full power of the functional programming language. See
Figure~\ref{fig:lex-sig-impl} for an example of implementing a
lexicalized version of the $\Sigma_{Synt}$ signature
from~\ref{sssec:example-sig} using the \texttt{ht->typer} abstraction.

\begin{figure}
  \centering
\begin{verbatim}
(ht->typer {{:head {:cat "n"}}      'N
            {:head {:cat "v"
                    :trans "true"}} (-> 'NP 'NP 'S)
            {:head {:cat "det"}}    (-> 'N (-> 'NP 'S) 'S)})
\end{verbatim}
  \caption{\label{fig:lex-sig-impl} A lexicalized version of the
    $\Sigma$(\_{Synt}) signature from~\ref{sssec:example-sig}.}
\end{figure}

In the above paragraphs, we have been referring to the constants that we
were defining as \emph{lexical constants}. That was to distinguish them
from the \emph{non-lexical constants} that we will introduce now. The
reason for having non-lexical constants is that not every constant of
every signature is linked to (generated by) some item of our lexical
database. Consider, for example, the empty string $\epsilon$ and the
string concatenation operator $+$ of the $\Sigma_{String}$ signature we
gave in~\ref{sssec:example-sig} or the quantifiers and logical
connectives of the $\Sigma_{Sem}$ signature in~\ref{sssec:acg-sem}.
These constants are not associated to any lexical entry or specifier and
therefore their representation differs. See
Figure~\ref{fig:nonlex-const} for an example of an non-lexicalized
constant from a semantic signature.

@@

Signatures of non-lexical constants can be simply defined by a mapping
from the constant names to their types as
in~\ref{fig:nonlex-sig-impl}. Such signatures will generally not be
sufficient by themselves and will need to be complemented by other
signatures which contain the lexical constants. Since we represent
signatures as unary relations, combining them to produce their unions or
intersections is just a matter of taking their disjunctions or
conjunctions, respectively. With this, we have enough to fully define
the other signatures of subsection~\ref{ssec:acg} in a lexicalized
manner, see Figure~\ref{fig:example-sig-impl}.

\begin{figure}
  \centering
\begin{verbatim}
(nonlex-sigr {'and?    (-> 'T 'T 'T)
              'imp?    (-> 'T 'T 'T)
              'forall? (-> (=> 'E 'T) 'T)
              'exists? (-> (=> 'E 'T) 'T)})
\end{verbatim}
  \caption{\label{fig:nonlex-sig-impl} A signature of non-lexical
    constants belonging to $\Sigma$(\_{Sem})
    from~\ref{sssec:acg-sem}. The single arrow \texttt{->} corresponds
    to linear implication while the double arrow \texttt{=>} corresponds
    to intuitionistic implication.}
\end{figure}

@@

(def sem-sig
  (ors (nonlex-sigr {'and?    (-> 'T 'T 'T)
                     'imp?    (-> 'T 'T 'T)
                     'forall? (-> (=> 'E 'T) 'T)
                     'exists? (-> (=> 'E 'T) 'T)})
       (ht->typer {{:head {:cat "n"}}      (-> 'E 'T)
                   {:head {:cat "v"
                           :trans "true"}} (-> 'E 'E 'T)})))
\end{verbatim}
  \caption{\label{fig:example-sig-impl} The definition of the
    lexicalized versions of the example signatures of
    subsection~\ref{ssec:acg}.}
\end{figure}

\subsubsection{Defining lexicons}

@@

In our relational framework, a mapping is most directly represented as a
binary relation, which is the representation that we have chosen. In the
previous subsubsection, we talked a lot about how individual entries in
the lexical database generate the items of a signature. Solving the
problem for signatures also solved it for lexicons, since their
structure is very similar: a signature is a set of pairs of constants
and types, a lexicon is a set of pairs of constants and
terms. Therefore, a lexical entry contributes the object terms for the
constants which it generated in the abstract signature.

@@

(def sem-lexo
  (with-sig-consts sem-sig
    (orr (lexicalizer sem-sig
                      (ht-lexiconr {{:head {:cat "n"}}
                                    ,(rt (ll [_ x] (_ x)))
                                    {:head {:cat "v"
                                            :trans "true"}}
                                    ,(rt (ll [_ s o] (_ s o)))}))
         (ht-lexiconr {{:head {:cat "det"
                               :lemma "un"}}
                       ,(rt (ll [P Q] (exists? (il [x] (and? (P x)
                                                             (Q x))))))
                       {:head {:cat "det"
                               :lemma "chaque"}}
                       ,(rt (ll [P Q] (forall? (il [x] (imp? (P x)
                                                             (Q x))))))}))))
\end{verbatim}
  \caption{\label{fig:lex-impl} The implementation of the
    $\mathcal{L}$(\_{Syntax}) and $\mathcal{L}$(\_{Sem}) lexicons from
    subsection~\ref{ssec:acg}.}
\end{figure}

@@

\subsection{Testing signatures and lexicons}

>>>>

\chapter{Wide-coverage abstract categorial grammars}

@@

\section{The tools and techniques}

@@

\subsection{The case for relational programming}

@@

Finally, some of the structures that we will encounter in our domain are
more directly modelled in mathematics as a relation instead of a
function. This has the benefit that to implement a relation in
relational programming, one just writes the relation, while if we were
forced to encode everything in functions, we would have to manage this
multiplicity of values manually. Examples of structures that have this
kind of structures if Frilex, our lexical database, where a single
wordform can be mapped to multiple hypertags. We will see another
example of a relation within our domain in the next section and we
will also discuss the advantages and drawbacks of modelling functions as
relations.

\subsection{Choice of implementation language}

@@

\section{Defining signatures and lexicons}

\subsection{Defining signatures}

In~\ref{ssec:sig}, we formally presented signatures as triples
$\mathopen{<}A, C, \tau\mathclose{>}$. The set $C$ can be easily
reconstructed as the domain of $\tau$ and the set $A$ is just the set of
all atomic types occurring in the range of $\tau$. Therefore, to define
the signature $\mathopen{<}A, C, \tau\mathclose{>}$, it is enough to
provide the function $\tau$, which in its set-theoretical realization is
just a set of pairs. In relational programming, a set can be easily
encoded as a unary relation that answers the membership test. Therefore,
our goal will be to construct to answer this relation and this relation
will be our final representation of a signature.

@@

\begin{description}
  \item[Function of wordform] We could imagine that every different
    wordform of the lexical database will end up being mapped to some
    constant in the signature. However, this is unsuitable since the
    lexical database is a relation that can assign more than one
    distinct hypertag to a single wordform (e.g. with different part of
    speech) and we would like to have constants of different types for
    each of these hypertags.
  \item[Function of hypertag] Conversely, we might imagine that the
    hypertags of the lexical database are mapped via a function to the
    constants of the signature. Here we run into trouble when we
    consider variations in spelling or phonology. We might have two
    items in our lexical database that share the same hypertag but whose
    wordforms still differ. We would like to have two distinct constants
    in our signature then, so that when these two constants are mapped
    by the surface lexicon, they will yield two different strings.
  \item[Function of lexical entry] Since neither the wordform nor the
    hypertag is enough, we could consider the entire lexical entry (an
    element of the lexical database, a pair of a wordform and one of its
    hypertags). This solves both of the above problems but it is still
    not perfect. Consider the case of a determiner and some semantic
    signature. Since our link between elements of the lexical database
    and of our signature is a function, there must be a constant in the
    signature for every item of our lexical database. If you recall
    in~\ref{ssec:acg-sem}, the constants of the semantic signature
    $\Sigma_{Sem}$ contained predicates for the common nouns $man$ and
    $woman$ and for the verb $loves$ and some common logical
    furniture. However, it did not contain any constants which would
    correspond to the determiners $every$ and $some$.
  \item[Partial function of lexical entry] We could fix the issue above
    by using a partial function. This accounts for cases when a lexical
    entry has no constants to represent it in a signature. Nevertheless,
    this only fixes half of the problem. We will consider the single
    lexical entry which represents e.g. the 3rd person singular present
    tense form $loves$ of the transitive verb $love$. A legitimate
    account of subject/object scope ambiguity would be to consider two
    constants in our abstract syntactic signature that have distinct
    images in some semantic lexicon, one giving the outer scope to the
    subject and the other to the object. This means that in order to
    define such a signature, we would need to link a single lexical
    entry in our database to more than one constant in the signature.
  \item[Relation between lexical entry and constant] Finally,
    considering all the above common scenarios, the only mathematical
    model general enough to handle all of the above cases is that of a
    relation. In our system, this will correspond exactly to a relation
    that will be provided by the grammar author.
\end{description}

@@

The system takes care of wiring up the relations and assembling the
signature. The grammar author's duty is to provide a 4-ary relation that
links the four parts of a lexical constant: the wordform, the hypertag,
the specifier and the type. This relation can be thought of as a
multi-valued ``function'' that receives as input a wordform and a
hypertag (that is, a single lexical entry) and outputs (possibly
multiple) pairs of specifiers and types. Since we are working in a
functional programming language that supports manipulation of
higher-order functions, we can provide helper functions which construct
these 4-ary relations for the grammar author in some simple cases. Our
library thus provides functions like \texttt{unitypedr}, which expects a
type and returns the 4-ary relation that assigns to every lexical item a
single constant whose specifier is $nil$ and whose type is the supplied
type, or \texttt{ht->typer}, which expects a mapping from hypertag
patterns to types and returns the 4-ary relation which tries to map the
hypertag against the patterns and yields constants with the
corresponding types and $nil$ specifiers. When other patterns emerge,
the grammar author is completely free to implement her new abstractions
using the full power of the functional programming language. See
Figure~\ref{fig:lex-sig-impl} for an example of implementing a
lexicalized version of the $\Sigma_{Synt}$ signature
from~\ref{ssec:example-sig} using the \texttt{ht->typer} abstraction.

\begin{figure}
  \centering
\begin{verbatim}
(ht->typer {{:head {:cat "n"}}      'N
            {:head {:cat "v"
                    :trans "true"}} (-> 'NP 'NP 'S)
            {:head {:cat "det"}}    (-> 'N (-> 'NP 'S) 'S)})
\end{verbatim}
  \caption{\label{fig:lex-sig-impl} A lexicalized version of the
    $\Sigma$(\_{Synt}) signature from~\ref{ssec:example-sig}.}
\end{figure}

In the above paragraphs, we have been referring to the constants that we
were defining as \emph{lexical constants}. That was to distinguish them
from the \emph{non-lexical constants} that we will introduce now. The
reason for having non-lexical constants is that not every constant of
every signature is linked to (generated by) some item of our lexical
database. Consider, for example, the empty string $\epsilon$ and the
string concatenation operator $+$ of the $\Sigma_{String}$ signature we
gave in~\ref{ssec:example-sig} or the quantifiers and logical
connectives of the $\Sigma_{Sem}$ signature in~\ref{ssec:acg-sem}.
These constants are not associated to any lexical entry or specifier and
therefore their representation differs. See
Figure~\ref{fig:nonlex-const} for an example of an non-lexicalized
constant from a semantic signature.

@@

Signatures of non-lexical constants can be simply defined by a mapping
from the constant names to their types as
in~\ref{fig:nonlex-sig-impl}. Such signatures will generally not be
sufficient by themselves and will need to be complemented by other
signatures which contain the lexical constants. Since we represent
signatures as unary relations, combining them to produce their unions or
intersections is just a matter of taking their disjunctions or
conjunctions, respectively. With this, we have enough to fully define
the other signatures of section~\ref{sec:acg} in a lexicalized
manner, see Figure~\ref{fig:example-sig-impl}.

\begin{figure}
  \centering
\begin{verbatim}
(nonlex-sigr {'and?    (-> 'T 'T 'T)
              'imp?    (-> 'T 'T 'T)
              'forall? (-> (=> 'E 'T) 'T)
              'exists? (-> (=> 'E 'T) 'T)})
\end{verbatim}
  \caption{\label{fig:nonlex-sig-impl} A signature of non-lexical
    constants belonging to $\Sigma$(\_{Sem})
    from~\ref{ssec:acg-sem}. The single arrow \texttt{->} corresponds
    to linear implication while the double arrow \texttt{=>} corresponds
    to intuitionistic implication.}
\end{figure}

@@

(def sem-sig
  (ors (nonlex-sigr {'and?    (-> 'T 'T 'T)
                     'imp?    (-> 'T 'T 'T)
                     'forall? (-> (=> 'E 'T) 'T)
                     'exists? (-> (=> 'E 'T) 'T)})
       (ht->typer {{:head {:cat "n"}}      (-> 'E 'T)
                   {:head {:cat "v"
                           :trans "true"}} (-> 'E 'E 'T)})))
\end{verbatim}
  \caption{\label{fig:example-sig-impl} The definition of the
    lexicalized versions of the example signatures of
    section~\ref{sec:acg}.}
\end{figure}

\subsection{Defining lexicons}

@@

In our relational framework, a mapping is most directly represented as a
binary relation, which is the representation that we have chosen. In the
previous subsection, we talked a lot about how individual entries in
the lexical database generate the items of a signature. Solving the
problem for signatures also solved it for lexicons, since their
structure is very similar: a signature is a set of pairs of constants
and types, a lexicon is a set of pairs of constants and
terms. Therefore, a lexical entry contributes the object terms for the
constants which it generated in the abstract signature.

@@

(def sem-lexo
  (with-sig-consts sem-sig
    (orr (lexicalizer sem-sig
                      (ht-lexiconr {{:head {:cat "n"}}
                                    ,(rt (ll [_ x] (_ x)))
                                    {:head {:cat "v"
                                            :trans "true"}}
                                    ,(rt (ll [_ s o] (_ s o)))}))
         (ht-lexiconr {{:head {:cat "det"
                               :lemma "un"}}
                       ,(rt (ll [P Q] (exists? (il [x] (and? (P x)
                                                             (Q x))))))
                       {:head {:cat "det"
                               :lemma "chaque"}}
                       ,(rt (ll [P Q] (forall? (il [x] (imp? (P x)
                                                             (Q x))))))}))))
\end{verbatim}
  \caption{\label{fig:lex-impl} The implementation of the
    $\mathcal{L}$(\_{Syntax}) and $\mathcal{L}$(\_{Sem}) lexicons from
    section~\ref{sec:acg}.}
\end{figure}

@@

\section{Testing signatures and lexicons}
